---
title: "Logic and types in R"
subtitle: "Lecture 02"
author: "Dr. Colin Rundel"
footer: "Sta 523 - Fall 2025"
format:
  revealjs:
    theme: slides.scss
    transition: fade
    slide-number: true
    self-contained: true
execute:
  echo: true
  warning: true
engine: knitr
---

```{r setup}
#| message: False
#| warning: False
#| include: False
options(
  htmltools.dir.version = FALSE, # for blogdown
  width=80
)
```

# Control Flow


## Conditional Control Flow

Conditional execution of code blocks is achieved via `if` statements. 

```{r}
x = c(1, 3)
```

. . .

:::: {.columns}
::: {.column width='50%'}
```{r}
if (3 %in% x) {
  print("Contains 3!")
}
```
:::

::: {.column width='50%'}
```{r}
if (1 %in% x)
  print("Contains 1!")
```
:::
::::

. . .

:::: {.columns}
::: {.column width='50%'}
```{r}
if (5 %in% x) {
  print("Contains 5!")
}
```
:::

::: {.column width='50%'}
```{r}
if (5 %in% x) {
  print("Contains 5!")
} else {
  print("Does not contain 5!")
}
```
:::
::::



## `if` is not vectorized

```{r}
x = c(1, 3)
```

. . .

```{r}
#| error: True
if (x == 1)
  print("x is 1!")
```

. . .

```{r}
#| error: True
if (x == 3)
  print("x is 3!")
```


::: {.aside}
Note that the behavior seen above (thrown errors) is new in R 4.2, previous versions will only throw warnings (using only the first value in the condition vector).
:::


## Collapsing logical vectors

There are a couple of helper functions for collapsing a logical vector down to a single value: `any`, `all`

```{r}
x = c(3,4,1)
```



:::: {.columns}
::: {.column width='50%'}
```{r}
x >= 2
any(x >= 2)
all(x >= 2)
```
:::

::: {.column width='50%'}
```{r}
x <= 4
any(x <= 4)
all(x <= 4)
```
:::
::::

. . .

```{r}
if (any(x == 3)) 
  print("x contains 3!")
```



## `else if` and `else`

:::: {.columns}
::: {.column width='50%'}
```{r}
x = 3

if (x < 0) {
  "x is negative"
} else if (x > 0) {
  "x is positive"
} else {
  "x is zero"
}
```
:::

::: {.column width='50%'}
```{r}
x = 0

if (x < 0) {
  "x is negative"
} else if (x > 0) {
  "x is positive"
} else {
  "x is zero"
}
```
:::
::::



## `if` return values

R's `if` conditional statements return a value (invisibly), the two following implementations are equivalent.

:::: {.columns}
::: {.column width='50%'}
```{r}
x = 5
```

```{r}
s = if (x %% 2 == 0) {
  x / 2
} else {
  3*x + 1
}
```

```{r}
s
```
:::

::: {.column width='50%'}
```{r}
x = 5
```

```{r}
if (x %% 2 == 0) {
  s = x / 2
} else {
  s = 3*x + 1
}
```

```{r}
s
```
:::
::::

::: {.aside}
Notice that conditional expressions are evaluated in the parent scope.
:::



## Exercise 2

Take a look at the following code below on the left, without running it in R what do you expect the outcome will be for each call on the right?

:::: {.columns}
::: {.column width='50%'}
```r
f = function(x) {
  # Check small prime
  if (x > 10 || x < -10) {
    stop("Input too big")
  } else if (x %in% c(2, 3, 5, 7)) {
    cat("Input is prime!\n")
  } else if (x %% 2 == 0) {
    cat("Input is even!\n")
  } else if (x %% 2 == 1) {
    cat("Input is odd!\n")
  }
}
```
:::

::: {.column width='50%'}
```r
f(1)
f(3)
f(8)
f(-1)
f(-3)
f(1:2)
f("0")
f("3")
f("zero")
```
:::
::::

::: {.aside}
More on functions next time
:::

```{r}
#| echo: False
countdown::countdown(minutes = 5)
```



## Conditionals and missing values

`NA`s can be particularly problematic for control flow,

:::: {.columns .small}
::: {.column width='50%'}
```{r}
#| error: True
if (2 != NA) {
  "Here"
}
```
:::

::: {.column width='50%'}
```{r}
#| error: True
2 != NA
```
:::
::::


. . .

:::: {.columns .small}
::: {.column width='50%'}

```{r}
#| error: True
if (all(c(1,2,NA,4) >= 1)) {
  "There"
}
```
:::

::: {.column width='50%'}

```{r}
#| error: True
all(c(1,2,NA,4) >= 1)
```
:::
::::

. . .

:::: {.columns .small}
::: {.column width='50%'}
```{r}
#| error: True
if (any(c(1,2,NA,4) >= 1)) {
  "There"
}
```
:::

::: {.column width='50%'}
```{r}
#| error: True
any(c(1,2,NA,4) >= 1)
```
:::
::::


## Testing for `NA`

To explicitly test if a value is missing it is necessary to use `is.na` (often along with `any` or `all`).

:::: {.columns}
::: {.column width='50%'}
```{r}
NA == NA
is.na(NA)
is.na(1)
```
:::

::: {.column width='50%'}
```{r}
is.na(c(1,2,3,NA))
any(is.na(c(1,2,3,NA)))
all(is.na(c(1,2,3,NA)))
```
:::
::::



